Features of Accretion-Phase Gravitational Wave Emission from Two-dimensional Rotating Core-Collapse Supernovae

### 2019-06-25 ms.tex
Text changes for the ArXiV (~3/4 done)

\title{Features of Accretion Phase Gravitational Wave Emission from Two-dimensional Rotating Core-Collapse Supernovae}

@@

% \abstract{}{}{}{}{} 
% 5 {} token are mandatory
 \begin{abstract}
  We explore the influence of progenitor mass and rotation on the gravitational wave (GW) emission from core collapse supernovae, during the post bounce, pre explosion, accretion phase.  We present the results from 15 two-dimensional (2D) neutrino radiation-hydrodynamic simulations from initial stellar collapse to $\sim$300 ms after core bounce.  We examine the features of the GW signals for four zero age main sequence (ZAMS) progenitor masses ranging from 12 \(M_\odot\) to 60 \(M_\odot\) and four core rotation rates from 0 rad s$^{-1}$ to 3 rad s$^{-1}$. We find that GW strain immediately around core bounce is fairly independent of ZAMS mass and---consistent with previous findings---that it is more heavily dependent on the core angular momentum.  At later times, all nonrotating progenitors exhibit loud GW emission, which we attribute to vibrational g-modes of the protoneutron star excited by convection in the post shock layer and the standing accretion shock instability (SASI).  We find that increasing rotation rates results in muting of the accretion phase GW signal due to centrifugal effects that inhibit convection in the post shock region, quench the SASI, and slow the rate at which the protoneutron star peak vibrational frequency increases.  Additionally, we verify the efficacy of our approximate general relativistic (GR) effective potential treatment of gravity by comparing our core bounce GW strains with the recent 2D GR results of other groups. 
   \keywords{gravitational waves --
                core-collapse --
                supernova
               }
\end{abstract}

@@

Core-collapse supernovae (CCSNe) became the first extra-solar multi-messenger object when SN 1987A was detected by the Kamiokande II experiment and Irvine-Michigan-Brookhaven (IMB) water Cherenkov detector in 1987 \citep{hirata:1987,bionta:1987} along with concurrent electromagnetic observations \citep[cf.][]{arnett:1989}. With the recent detection of a neutron star merger--GW170817--in both photons and gravitational waves (GWs) by the LIGO and Virgo collaborations \citep{abbott:2016} we have entered the era of {\it GW} multimessenger astronomy.  
So far, only the mergers of black hole binaries and a neutron star binary have been detected in GWs, but CCSNe are also predicted to be prodigious GW sources, though not quite as ``loud'' as compact object binary mergers.
Accurate predictions of the expected GW signal from CCSNe is key to increasing the likelihood of detection by GW observatories such as Advanced LIGO (aLIGO) and Advanced Virgo (AdV) and will be crucial in our ability to extract physical meaning from a future CCSN GW detection \citep{abdik:2014,gossan:2016}.

CCSNe are routinely observed in the electromagnetic (EM) window, and the data collecting power of synoptic surveys such as the Large Synoptic Survey Telescope (LSST) and Zwicky Transient Facility (ZTF) may increase the volume of such data for CCSNe by orders of magnitude \citep{ivezic:2008,bellm:2019}.
Still, until the late nebular phase, which is often too dim to be easily observed for distant CCSNe, the EM emission arises from the very outermost layers of the progenitor star and the central core regions, where the explosion is driven, are obscured. 
This makes it challenging to connect EM emission from CCSNe directly to the mechanism that powers them.
Due to their relatively small interaction probabilities with matter, both neutrinos and GWs offer windows through which to peer directly into the heart of a CCSN explosion.  
Moreover, these observations have broader astrophysical applications: restricting nuclear equations of state, verifying angular momentum transport in plasmas, and better understanding stellar rotation.

An observation of either GWs or neutrino emission from a nearby CCSN combined with multiband observations would allow us to place unique constraints on the physics of the explosion mechanism and key nuclear physics, such as the nuclear equation of state.  
There has yet to be a single astrophysical object detected via all three of these messengers.  
Albeit a rare event, a Galactic CCSN offers the perfect opportunity to observe such a multimessenger ``trifecta.''  
In order to increase our chances of ``hearing'' such an event in GWs, and in order to be able to extract the greatest scientific meaning from them, we need accurate predictions for CCSN GW signals from the wide range of initial conditions that give rise to these stellar explosions.

Modeling GWs from CCSNe incurs all of the challenges of simulating the CCSN mechanism itself, along with a heightened emphasis on the importance of the general relativistic (GR) treatment of gravity. 
The increased expense of including a fully dynamical spacetime evolution coupled to general relativistic dynamics \citep[cf.][]{ott:2009, ott:2012} can further reduce the size of the parameter space that is feasible to explore. 
Approximations that maintain sufficient numerical accuracy become necessary in order to reduce computational cost.  
A common approach for CCSNe, particularly in 2D, is the conformal flatness condition (CFC) approximation wherein the spatial three-metric is obtained approximately from the flat spacetime three-metric.
CFC has been shown to accurately reproduce prebounce and early post bounce signals from CCSNe to within a few percent when compared to direct solutions to Einstein's field equations  \citep{ott:2007}.  \textcolor{red}{[6] Likewise, while some differences appear, \citet{shibata:2004} find good qualitative agreement between the effective GR potential and CFC.  [3] This conformal flatness approach has also been extended to an ``augmented CFC" (xCFC) scheme as introduced by \citet{saijo:2004}, refined by \citet{cordero-carrion:2009}, and utilized by \citet{bmuller:2019}.}
A further approximation, also common in simulations of the CCSN mechanism, is to couple an effective GR gravitational potential to otherwise Newtonian dynamics \citep{rampp:2002, marek:2006, bruenn:2016, oconnor:2018, morozova:2018}.  
This relativistic effective potential empirically satisfies the solution to hydrostatic equilibrium according to a modified Tolman-Oppenheimer-Volkoff equation (TOV) \citep{rampp:2002, marek:2006}.
This approach further reduces the computational expense of CCSN simulations relative to the CFC approach and reproduces fairly accurately gross features of CCSN simulations \citep{marek:2006, muller:2012,oconnor:2018}.

After the infalling matter from collapse reaches nuclear densities, the core nuclei dissolve into nucleons and, eventually, the strong force becomes repulsive, halting the material infall.  On the time scale of tens of microseconds, the subsonic inner core encounters the supersonic outer core, forming a shock front.  As this shock front photodissociates overlying material and releases an enormous neutrino flux, it leaves behind a negative entropy gradient \citep{mazurek:1982,bruenn:1985,bruenn:1989}.  This scenario is unstable according to the Ledoux criterion, causing prompt convection in the post shock region \citep{burrows:1992}, therefore creating an associated emission of gravitational radiation \citep{marek:2009b,ott:2009}.  This prompt convection is an important feature that occurs in simulations that incorporate either GR or Newtonian treatments of gravity, during the early post bounce phase \citep{muller:2017,richers:2017,nagakura:2018}.

Early research into GW emission from CCSNe focused on the bounce and early post bounce phase of the explosion in rotating progenitors.
These investigations found that increasing the angular momentum of the core leads to a larger strain peak at bounce \citep{muller:1982,moench:1991,yamada:1995,zwerger:1997,dimm:2002,kotake:2003,shibata:2004}.  
More recent investigations of rotating core collapse examine the role of the angular momentum distribution within the supernova progenitor and find it only important in the rapid rotation regime, where the ratio of kinetic to gravitational potential energy ($T/|W|$) $ \gtrsim 8\%$ at bounce \citep{abdik:2014}. In order to examine GW emission at later times, different groups have considered other factors for nonrotating cases---for example, convection in the post shock region \citep{burrows:1996,muller:1997,muller:2004,murphy:2009,marek:2009b}, the standing accretion shock instability (SASI) \citep{blondin:2003,blondin:2006,ohnishi:2006,foglizzo:2007,scheck:2008,iwakami:2009,fernandez:2010}\textcolor{red}{[9], and protoneutron star (PNS) vibrational modes \citep{cerda-duran:2013,torres-forne:2018,torres-forne:2019}}.  \citet{moro:2018} investigate GW emission for moderate rotational speeds ($\Omega_{\mathrm{core}} = 0.2$ rad s$^{-1}$) for a single progenitor mass ($13 M_\odot$) over one second post bounce.  \citet{pan:2018}, \citet{kuroda:2018}, \textcolor{red}{[3]}\citet{cerda-duran:2013}, and \citet{ott:2011} investigate the relationship between black hole formation and GW emission, for a nonrotating 40 $M_\odot$, nonrotating  70 $M_\odot$, \textcolor{red}{[3] a rotating 35 $M_\odot$,} and a rotating 75 $M_\odot$ progenitor, respectively.  These works also find stronger GW emission at bounce with increased progenitor angular momentum and loud GW emission at later times for nonrotating CCSNe.

In this work, we present 15 axisymmetric (2D) neutrino radiation-hydrodynamic CCSN simulations.  
Our parameter space spans four progenitor masses ranging from $12\,M_\odot-60\,M_\odot$ \citep{Suk:2016} and four peak core rotation speeds: $0-3 \text{ rad s}^{-1}$.  
We examine the variation in key features of the GW emission from CCSNe at these different masses and rotation rates.
Rapid rotation rates up to 2 and 3 rad s$^{-1}$ are likely rare in typical massive stars at solar metallicity due to efficient transport and loss of angular momentum \citep{heger:2005}.
While \citet{woosley:2006} observe that only 1\% of massive stars may reach the rapid rotation regime, there are high uncertainties in the stellar mass loss and magnetic braking calculations \citep{smith:2014}.  \textcolor{red}{[5] Moreover, albeit a small percentage, \citet{demink:2013} show the distinct possibility of rapidly rotating stars formed from binary interactions.} Thus, there is some likelihood of rapidly rotating supernova progenitors in the mass range we explore.

%However, when placed in the context of gravitational wave observations, we expect magnetic breaking due to the Taylor-Spruit Dynamo to slow the progenitor rotation \citep{spruit:2002}.

%place in context of rotation
%TS dynamo breaks core -->0-1 regime
%GW observers should look here
%Drop in low frequencies between 0->1 rad/s

In addition to the breadth of parameter space we cover, we also explore the role of rotation up to 300 ms post bounce.  
We find that rotation restricts the growth of SASI by centrifugally flattening the shock, leaving it slightly oblate. Likewise, the positive angular momentum gradient created by the rotation stabilizes the post shock convection according to the Solberg-H{\o}iland stability criterion \citep{endal:1978,fryer:2000}.  Not only are the SASI and post shock convection contributions to the gravitational radiation diminished, but the PNS vibrational signals are damped because of less turbulent downflows of matter onto the PNS surface.  
This results in a ``muting'' of the GW signal with increasing rotation speeds.
While the origins of this muting are physical, such behavior may not be seen in full 3D simulations of CCSNe due to the appearance of spiral modes of the SASI \textcolor{red}{[4] and magnetorotational instabilities (MRI)} \citep{andresen:2018,cerda-duran:2007} .

\textcolor{red}{[1] Compared to previous works, the strength of this project is its ability analyze GWs from multiple progenitors hundreds of milliseconds post bounce while accurately accounting for rotation and neutrinos.  The wide breadth of parameter space we examine allows us to reveal certain rotational effects on the GW signal in the context of a controlled study.}

@@

This paper is organized as follows:  In Section \ref{sec:method} we present our methods and treatment of microphysics within our \texttt{FLASH} simulations.  We present a new method to applying initial rotation to the progenitor.  Because each progenitor evolves at a different rate and we terminate our simulations at 300 ms post bounce, we refrain from asserting which explode.  Rather, in Section \ref{sec:results}, we begin by addressing the shock front evolution for each of the progenitor masses and initial rotation velocities.  We then verify our gravitational treatment by comparing our bounce signal to general relativistic simulations.  We explore the effect of rotation on GWs emitted hundreds of milliseconds after core bounce and discuss implications on their detectability.  In Section \ref{sec:summary} we conclude and summarize the influence of rotation on GWs from initial collapse to 300 ms post bounce.

@@

\section{Methods and Simulation Setup}
\label{sec:method}
We utilize the \texttt{FLASH} (version 4) multiscale, multiphysics adaptive mesh refinement (AMR) simulation framework for our simulations \citep{fryxell:2000,dubey:2009}.\footnote{http://flash.uchicago.edu/site/}  We employ a modified, general relativistic, effective potential \citep{marek:2006, oconnor:2018} incorporated into the multipole Poisson solver of \citet{couch:2013a}, where we retain spherical harmonic orders up through 16.   We utilize the SFHo equation of state (EOS) in all of our 15 simulations \citep{steiner:2013}.  Our grid setup is a 2D cylindrical geometry with the PARAMESH (v.4-dev) library for AMR  \citep{macneice:2000}.  The outer boundary is $10^4$ km in all directions, with nine levels of refinement, yielding a finest grid spacing of about 0.65 km.
The maximum allowed level of refinement is decreased as a function of spherical radius, $r$, in order to maintain a resolution aspect ratio, $\Delta x_i / r$, of about 0.01, corresponding approximately to an ``angular'' resolution of 0.5 degrees.

%M1 section
Neutrinos play a vital role in core-collapse supernovae.  Directly after collapse, they provide an avenue through which the PNS can cool.  As the shock propagates outwards, they also provide heating in the gain region that is crucial in reviving the explosion, according to the neutrino heating mechanism.  The opacity of the material to these outflowing neutrinos must be carefully accounted for in an energy dependent way.  We incorporate a multidimensional, multispecies, energy dependent, two moment scheme with an analytic closure, or the so called M1 scheme.  Our implementation is based on \citet{oconnor:2015}, \citet{shibata:2011}, and \citet{cardall:2013}.  A detailed outline of the M1 implementation in \texttt{FLASH} is in \citet{oconnor:2018}.  \textcolor{red}{[2] This combination of rotation and the M1 neutrino treatment is similar to the work of \citet{obergaulinger:2017} and \citet{obergaulinger:2018}.}  In order to reduce computational cost to explore the wide parameter space for our study, we neglect velocity dependent neutrino transport and do not account for inelastic neutrino-electron scattering.
We use 12 energy bins spaced logarithmically up to 250 MeV, and the full set of rates and opacities we use is described in \citet{oconnor:2017a}. 
Specifically, we use the effective, many body, corrected rates
for neutrino-nucleon, neutral current scattering of \citet{horowitz:2017}.

We use the 12, 20, 40, and 60 \Msun non-rotating, solar-metallicity progenitors models from \citet{Suk:2016} for the present work.

\begin{figure}[t]
    \centering
    \includegraphics[scale=0.45]{omega_vs_r.pdf}
    \caption{The rotation profiles for five of the \citet{heger:2005} progenitors.  Each solid line represents \citeauthor{heger:2005}'s (\citeyear{heger:2005}) model and the dashed lines are Equation (\ref{eq:omega}) applied to the respective progenitor, with the appropriate differential rotation parameter.  }
    \label{fig:ovsr}
\end{figure}

@@

The progenitor models we use are evolved without rotation.
At the start of core collapse, when we map the 1D models into our 2D grid, we apply an artificial rotation profile,
\begin{equation}
    \Omega(r) = \Omega_0 \bigg[1 + \bigg(\frac{r}{A}\bigg)^2 \bigg]^{-1}, 
    \label{eq:omega}
\end{equation}

@@

The precise rotation rates and profiles of massive stellar cores at collapse is uncertain.
Previous work \citep[e.g.,][]{abdik:2014} treated the differential rotation parameter $A$ as a free parameter and explored the impact of its variation.
Examining the stellar evolution models of \citet{heger:2005}, which include angular momentum transport due to the Tayler-Spruit dynamo \citep{spruit:2002}, we find that $A$ is strongly determined by the {\it compactness} \citep{oconnor:2011} of the stellar core.
In order to demonstrate this, we fit the rotation profiles of the 20 progenitor models from \citet{heger:2005} to Equation (\ref{eq:omega}) in order to determine the best fit $A$.
The models of \citet{heger:2005} include stars of ZAMS masses 12, 15, 20, 25, and 35 $M_{\odot}$, with various angular momentum transport parameters and initial ZAMS rotation rates.  
Using the \texttt{curve\_fit} function (in the \texttt{scipy.optimize} library) available in \texttt{Python}, we obtain $A$ values that correspond to the most accurate fits of Equation (\ref{eq:omega}) to the rotation profiles of these models.  Figure \ref{fig:ovsr} displays the radial, rotation profile for five of the aforementioned progenitors, compared to our implementation of Equation (\ref{eq:omega}), with the best fit $A$ value.

The core compactness as introduced by \citet{oconnor:2011} is defined as,
\begin{equation}
    \xi_M = \left.\frac{M/M_{\odot}}{R(M_\mathrm{bary}=M)/1000 \text{km}}\right\vert_\mathrm{collapse} ,
\end{equation} 
where we choose $M = 2.5 M_\odot$, and $R(M_{\mathrm{bary}}=M) $ as the radius at which the internal baryonic mass is $2.5M_\odot$, at collapse.  
Figure \ref{fig:a_vs_comp} shows the compactness parameters from the \citet{heger:2005} models (blue stars) plotted against their best fit $A$ values.
A clear linear relation exists between $A$ and $\xi_{2.5}$.

@@

\begin{table}[t]
\begin{tabular}{c|c|c}
Progenitor Mass [$M_\odot$] & Compactness & \textit{A}[1000 km] \\
\hline
12  & 0.0738 &         0.8123             \\
20  & 0.2785 &         1.021            \\
40  & 0.5341 &         1.282           \\
60  & 0.1708 &         0.9112          
\end{tabular}
\caption{Listed values for ZAMS mass, compactness calculated from \citet{Suk:2016}, and differential rotation parameter \textit{A}.}
\label{table:compact}
\end{table}

Using this relationship, we calculate optimal $A$ values for the four \citet{Suk:2016} progenitors we use in this work (orange circles in Figure \ref{fig:a_vs_comp}).  
The full list of progenitor masses, compactness values, and A's we use is given in Table ~\ref{table:compact}.

>>>>

\title{Features of Accretion-Phase Gravitational Wave Emission from Two-dimensional Rotating Core-Collapse Supernovae}

@@

% \abstract{}{}{}{}{} 
% 5 {} token are mandatory
 \begin{abstract}
  We explore the influence of progenitor mass and rotation on the gravitational-wave (GW) emission from core-collapse supernovae, during the postbounce, preexplosion, accretion-phase.  We present the results from 15 two-dimensional (2D) neutrino radiation-hydrodynamic simulations from initial stellar collapse to $\sim$300 ms after core bounce.  We examine the features of the GW signals for four zero-age main sequence (ZAMS) progenitor masses ranging from 12 \(M_\odot\) to 60 \(M_\odot\) and four core rotation rates from 0 to 3 rad s$^{-1}$. We find that GW strain immediately around core bounce is fairly independent of ZAMS mass and---consistent with previous findings---that it is more heavily dependent on the core angular momentum.  At later times, all nonrotating progenitors exhibit loud GW emission, which we attribute to vibrational g-modes of the protoneutron (PNS) star excited by convection in the postshock layer and the standing accretion shock instability (SASI).  We find that increasing rotation rates results in muting of the accretion-phase GW signal due to centrifugal effects that inhibit convection in the postshock region, quench the SASI, and slow the rate at which the PNS star peak vibrational frequency increases.  Additionally, we verify the efficacy of our approximate general relativistic (GR) effective potential treatment of gravity by comparing our core bounce GW strains with the recent 2D GR results of other groups. 
   \keywords{supernovae: general --
                core-collapse --
                stars: massive
               }
\end{abstract}

@@

Core-collapse supernovae (CCSNe) became the first extra-solar multimessenger objects when SN 1987A was detected by the Kamiokande II experiment and Irvine-Michigan-Brookhaven water Cerenkov detector in 1987 \citep{bionta:1987,hirata:1987} along with concurrent electromagnetic (EM) observations \citep[see][]{arnett:1989}. With the recent detection of a neutron star merger--GW170817--in both photons and gravitational waves (GWs) by the LIGO and Virgo collaborations \citep{abbott:2016}, we have entered the era of GW multimessenger astronomy.  
So far, only the mergers of black hole binaries and a neutron star binary have been detected in GWs, but CCSNe are also predicted to be prodigious GW sources, although not quite as ``loud'' as compact object binary mergers.
Accurate predictions of the expected GW signal from CCSNe is key to increasing the likelihood of detection by GW observatories such as Advanced LIGO (aLIGO) and Advanced Virgo (AdV) and will be crucial in our ability to extract physical meaning from a future CCSN GW detection \citep{abdik:2014,gossan:2016}.

CCSNe are routinely observed in the EM window, and the data-collecting power of synoptic surveys such as the Large Synoptic Survey Telescope and Zwicky Transient Facility may increase the volume of such data for CCSNe by orders of magnitude \citep{ivezic:2008,bellm:2019}.
Still, until the late nebular phase, which is often too dim to be easily observed for distant CCSNe, the EM emission arises from the very outermost layers of the progenitor star and the central core regions, where the explosion is driven, are obscured. 
This makes it challenging to connect EM emission from CCSNe directly to the mechanism that powers them.
Due to their relatively small interaction probabilities with matter, both neutrinos and GWs offer windows through which to peer directly into the heart of a CCSN explosion.  
Moreover, these observations have broader astrophysical applications: restricting nuclear equations of state, verifying angular momentum transport in plasmas, and better understanding stellar rotation.

An observation of either GWs or neutrino emission from a nearby CCSN combined with multiband observations would allow us to place unique constraints on the physics of the explosion mechanism and key nuclear physics, such as the nuclear equation of state (EOS).  
There has yet to be a single astrophysical object detected via all three of these messengers.  
Albeit a rare event, a Galactic CCSN offers the perfect opportunity to observe such a multimessenger ``trifecta.''  
In order to increase our chances of ``hearing'' such an event in GWs, and in order to be able to extract the greatest scientific meaning from them, we need accurate predictions for CCSN GW signals from the wide range of initial conditions that give rise to these stellar explosions.

Modeling GWs from CCSNe incurs all of the challenges of simulating the CCSN mechanism itself, along with a heightened emphasis on the importance of the general relativistic (GR) treatment of gravity. 
The increased expense of including a fully dynamical spacetime evolution coupled to GR dynamics \citep[see][]{ott:2009, ott:2012} can further reduce the size of the parameter space that it is feasible to explore. 
Approximations that maintain sufficient numerical accuracy become necessary in order to reduce computational cost.  
A common approach for CCSNe, particularly in 2D, is the conformal flatness condition (CFC) approximation wherein the spatial three-metric is obtained approximately from the flat spacetime three-metric.
CFC has been shown to accurately reproduce prebounce and early postbounce signals from CCSNe to within a few percent when compared with direct solutions to Einstein's field equations  \citep{ott:2007}.  Likewise, while some differences appear, \citet{shibata:2004} find good qualitative agreement between the effective GR potential and CFC.  This conformal flatness approach has also been extended to an ``augmented CFC" scheme as introduced by \citet{saijo:2004}, refined by \citet{cordero-carrion:2009}, and utilized by \citet{bmuller:2019}.
A further approximation, also common in simulations of the CCSN mechanism, is to couple an effective GR gravitational potential to otherwise Newtonian dynamics \citep{rampp:2002, marek:2006, bruenn:2016,  morozova:2018, oconnor:2018}.  
This relativistic effective potential empirically satisfies the solution to hydrostatic equilibrium according to a modified Tolman-Oppenheimer-Volkoff equation \citep{rampp:2002, marek:2006}.
This approach further reduces the computational expense of CCSN simulations relative to the CFC approach and reproduces fairly accurately gross features of CCSN simulations \citep{marek:2006, muller:2012,oconnor:2018}.

After the infalling matter from collapse reaches nuclear densities, the core nuclei dissolve into nucleons and, eventually, the strong force becomes repulsive, halting the material infall.  On the time scale of tens of microseconds, the subsonic inner core encounters the supersonic outer core, forming a shock front.  As this shock front photodissociates overlying material and releases an enormous neutrino flux, it leaves behind a negative entropy gradient \citep{mazurek:1982,bruenn:1985,bruenn:1989}.  This scenario is unstable according to the Ledoux criterion, causing prompt convection in the postshock region \citep{burrows:1992}, therefore creating an associated emission of gravitational radiation \citep{marek:2009b,ott:2009}.  This prompt convection is an important feature that occurs in simulations that incorporate either GR or Newtonian treatments of gravity during the early postbounce phase \citep{muller:2017,richers:2017,nagakura:2018}.

Early research into GW emission from CCSNe focused on the bounce and early postbounce phase of the explosion in rotating progenitors.
These investigations found that increasing the angular momentum of the core leads to a larger strain peak at bounce \citep{muller:1982,moench:1991,yamada:1995,zwerger:1997,dimm:2002,kotake:2003,shibata:2004}.  
More recent investigations of rotating core collapse examine the role of the angular momentum distribution within the supernova progenitor and find it is only important in the rapid rotation regime, where the ratio of kinetic to gravitational potential energy ($T/|W|$) $ \gtrsim 8\%$ at bounce \citep{abdik:2014}. In order to examine GW emission at later times, different groups have considered other factors for nonrotating cases---for example, convection in the postshock region \citep{burrows:1996,muller:1997,muller:2004,marek:2009b,murphy:2009}, the standing accretion shock instability (SASI) \citep{blondin:2003,blondin:2006,ohnishi:2006,foglizzo:2007,scheck:2008,iwakami:2009,fernandez:2010}, and protoneutron star (PNS) vibrational modes \citep{cerda-duran:2013,torres-forne:2018,torres-forne:2019}.  \citet{moro:2018} investigate GW emission for moderate rotational speeds ($\Omega_{\mathrm{core}} = 0.2$ rad s$^{-1}$) for a single progenitor mass ($13 M_\odot$) over 1 second postbounce.  \citet{pan:2018}, \citet{kuroda:2018}, \citet{cerda-duran:2013}, and \citet{ott:2011} investigate the relationship between black hole formation and GW emission, for a nonrotating 40 $M_\odot$, a nonrotating  70 $M_\odot$, a rotating 35 $M_\odot$, and a rotating 75 $M_\odot$ progenitor, respectively.  These studies also find stronger GW emission at bounce with increased progenitor angular momentum and loud GW emission at later times for nonrotating CCSNe.

In this work, we present 15 axisymmetric (2D) neutrino radiation-hydrodynamic CCSN simulations.  
Our parameter space spans four progenitor masses ranging from $12\,M_\odot$ to $60\,M_\odot$ \citep{Suk:2016} and four peak core rotation speeds: $0-3 \text{ rad s}^{-1}$.  
We examine the variation in key features of the GW emission from CCSNe at these different masses and rotation rates.
Rapid rotation rates up to 2 and 3 rad s$^{-1}$ are likely rare in typical massive stars at solar metallicity due to efficient transport and loss of angular momentum \citep{heger:2005}.
While \citet{woosley:2006} observe that only 1\% of massive stars may reach the rapid rotation regime, there are high uncertainties in the stellar mass loss and magnetic braking calculations \citep{smith:2014}.  Moreover, albeit a small percentage, \citet{demink:2013} show the distinct possibility of rapidly rotating stars formed from binary interactions. Thus, there is some likelihood of rapidly rotating supernova progenitors in the mass range we explore.

In addition to the breadth of parameter space we cover, we also explore the role of rotation up to 300 ms postbounce.  
We find that rotation restricts the growth of SASI by centrifugally flattening the shock, leaving it slightly oblate. Likewise, the positive angular momentum gradient created by the rotation stabilizes the postshock convection according to the Solberg-H{\o}iland stability criterion \citep{endal:1978,fryer:2000}.  Not only are the SASI and postshock convection contributions to the gravitational radiation diminished, but the PNS vibrational signals are damped because of less turbulent downflows of matter onto the PNS surface.  
This results in a ``muting'' of the GW signal with increasing rotation speeds.
While the origins of this muting are physical, such behavior may not be seen in full 3D simulations of CCSNe due to the appearance of spiral modes of the SASI and magnetorotational instabilities (MRI) \citep{cerda-duran:2007,andresen:2018} .

Compared with previous works, the strength of this project is its ability analyze GWs from multiple progenitors hundreds of milliseconds postbounce while accurately accounting for rotation and neutrinos.  The wide breadth of parameter space we examine allows us to reveal certain rotational effects on the GW signal in the context of a controlled study.

@@

This paper is organized as follows:  in Section \ref{sec:method} we present our methods and treatment of microphysics within our \texttt{FLASH} simulations.  We present a new method for applying initial rotation to the progenitor.  Because each progenitor evolves at a different rate and we terminate our simulations at 300 ms postbounce, we refrain from asserting which explode.  Rather, in Section \ref{sec:results}, we begin by addressing the shock front evolution for each of the progenitor masses and initial rotation velocities.  We then verify our gravitational treatment by comparing our bounce signal to GR simulations.  We explore the effect of rotation on GWs emitted hundreds of milliseconds after core bounce and discuss implications on their detectability.  In Section \ref{sec:summary} we conclude and summarize the influence of rotation on GWs from initial collapse to 300 ms postbounce.

@@

\section{Methods and Simulation Setup}
\label{sec:method}
We utilize the \texttt{FLASH} (version 4) multiscale, multiphysics adaptive mesh refinement simulation framework for our simulations \citep{fryxell:2000,dubey:2009}.\footnote{http://flash.uchicago.edu/site/}  We employ a modified, GR, effective potential \citep{marek:2006, oconnor:2018} incorporated into the multipole Poisson solver of \citet{couch:2013a}, where we retain spherical harmonic orders up through 16.   We utilize the SFHo EOS in all of our 15 simulations \citep{steiner:2013}.  Our grid setup is a 2D cylindrical geometry with the PARAMESH (v.4-dev) library for adaptive mesh refinement  \citep{macneice:2000}.  The outer boundary is $10^4$ km in all directions, with nine levels of refinement, yielding a finest grid spacing of about 0.65 km.
The maximum allowed level of refinement is decreased as a function of spherical radius, $r$, in order to maintain a resolution aspect ratio, $\Delta x_i / r$, of about 0.01, corresponding approximately to an ``angular'' resolution of $0.5^{\circ}$ .

%M1 section
Neutrinos play a vital role in CCSNe.  Directly after collapse, they provide an avenue through which the PNS can cool.  As the shock propagates outward, they also provide heating in the gain region that is crucial in reviving the explosion, according to the neutrino heating mechanism.  The opacity of the material to these outflowing neutrinos must be carefully accounted for in an energy-dependent way.  We incorporate a multidimensional, multispecies, energy-dependent, two-moment scheme with an analytic closure, or the so-called M1 scheme.  Our implementation is based on \citet{oconnor:2015}, \citet{shibata:2011}, and \citet{cardall:2013}.  A detailed outline of the M1 implementation in \texttt{FLASH} is in \citet{oconnor:2018}.  This combination of rotation and the M1 neutrino treatment is similar to the work of \citet{obergaulinger:2017} and \citet{obergaulinger:2018}.  In order to reduce computational costs to explore the wide parameter space for our study, we neglect velocity-dependent neutrino transport and do not account for inelastic neutrino-electron scattering.
We use 12 energy bins spaced logarithmically up to 250 MeV, and the full set of rates and opacities we use is described in \citet{oconnor:2017a}. 
Specifically, we use the effective, many-body, corrected rates
for neutrino-nucleon, neutral current scattering of \citet{horowitz:2017}.

We use the 12, 20, 40, and 60 \Msun nonrotating, solar-metallicity progenitors models from \citet{Suk:2016} for the present work.

\begin{figure}[t]
    \centering
    \includegraphics[scale=0.45]{omega_vs_r.pdf}
    \caption{The rotation profiles for five of the \citet{heger:2005} progenitors.  Each solid line represents \citeauthor{heger:2005}'s (\citeyear{heger:2005}) model, and the dashed lines are Equation (\ref{eq:omega}) applied to the respective progenitor, with the appropriate differential rotation parameter.  }
    \label{fig:ovsr}
\end{figure}

@@

The progenitor models we use are evolved without rotation.
At the start of core collapse, when we map the 1D models into our 2D grid, we apply an artificial rotation profile
\begin{equation}
    \Omega(r) = \Omega_0 \bigg[1 + \bigg(\frac{r}{A}\bigg)^2 \bigg]^{-1}, 
    \label{eq:omega}
\end{equation}

@@

The precise rotation rates and profiles of massive stellar cores at collapse are uncertain.
Previous work \citep[e.g.,][]{abdik:2014} treated the differential rotation parameter $A$ as a free parameter and explored the impact of its variation.
Examining the stellar evolution models of \citet{heger:2005}, which include angular momentum transport due to the Tayler-Spruit dynamo \citep{spruit:2002}, we find that $A$ is strongly determined by the {\it compactness} \citep{oconnor:2011} of the stellar core.
In order to demonstrate this, we fit the rotation profiles of the 20 progenitor models from \citet{heger:2005} to Equation (\ref{eq:omega}) in order to determine the best-fit $A$.
The models of \citet{heger:2005} include stars of zero-age main sequence (ZAMS) masses 12, 15, 20, 25, and 35 $M_{\odot}$, with various angular momentum transport parameters and initial ZAMS rotation rates.  
Using the \texttt{curve\_fit} function (in the \texttt{scipy.optimize} library) available in \texttt{Python}, we obtain $A$ values that correspond to the most accurate fits of Equation (\ref{eq:omega}) to the rotation profiles of these models.  Figure \ref{fig:ovsr} displays the radial, rotation profile for five of the aforementioned progenitors, compared with our implementation of Equation (\ref{eq:omega}), with the best-fit $A$ value.

The core compactness as introduced by \citet{oconnor:2011} is defined as
\begin{equation}
    \xi_M = \left.\frac{M/M_{\odot}}{R(M_\mathrm{bary}=M)/1000 \text{km}}\right\vert_\mathrm{collapse} ,
\end{equation} 
where we choose $M = 2.5 M_\odot$, and $R(M_{\mathrm{bary}}=M) $ as the radius at which the internal baryonic mass is $2.5M_\odot$, at collapse.  
Figure \ref{fig:a_vs_comp} shows the compactness parameters from the \citet{heger:2005} models (blue stars) plotted against their best-fit $A$ values.
A clear linear relation exists between $A$ and $\xi_{2.5}$.

@@

\begin{table}[t]
\begin{tabular}{c|c|c}
Progenitor Mass ($M_\odot$) & Compactness & \textit{A}(1000 km) \\
\hline
12  & 0.0738 &         0.8123             \\
20  & 0.2785 &         1.021            \\
40  & 0.5341 &         1.282           \\
60  & 0.1708 &         0.9112          
\end{tabular}
\caption{Listed values for ZAMS mass, Compactness Calculated from \citet{Suk:2016}, and Differential Rotation Parameter \textit{A}.}
\label{table:compact}
\end{table}

Using this relationship, we calculate optimal $A$ values for the four \citet{Suk:2016} progenitors we use in this work (orange circles in Figure \ref{fig:a_vs_comp}).  
The full list of progenitor masses, compactness values, and \textit{A} values we use is given in Table ~\ref{table:compact}.

As a note, we choose to omit the 40 $M_\odot$ progenitor at $\Omega_0 = 3$ rad s$^{-1}$ from our following analysis with a numerically motivated rationale.  The $\xi_{2.5}$ value of this progenitor is nearly double that of the 20 $M_\odot$ progenitor (the next closest compactness value).  This fact displays that the 40 $M_\odot$ has a much larger differential rotation parameter compared with the other progenitors, resulting in a nearly solid-body rotation of the core. This endows the core in the 40 \Msun model with drastically more angular momentum than the other models.  The vast amount of angular momentum ultimately led to numerical instabilities in our calculations; thus, we omit the 40 $M_\odot$, with $\Omega_0 = 3$ rad s$^{-1}$ from our analysis.  \\
%So much so, in fact, core bounce is significantly delayed and occurs at much lower central densities due to the strong centrifugal barrier.
%We outline in greater detail the resulting distortion of the GW signal in Section \ref{sec:results}.  Thus, in order to preserve a maximum upper limit on the angular momentum (within a 1.75 $M_\odot$ mass radius) of $\sim 2.4\times 10^{49}$ erg s, we omit the 40 $M_\odot$, with $\Omega_0 = 3$ rad s$^{-1}$ from our analysis.  \\
